---
title: "Assignment 2 - FMLA"
author: "Deepika"
date: "2026-10-03"
output:
  pdf_document: default
  html_document: default
---
```{r}
# Load required packages
library(class)
library(caret)
library(dplyr)
```

#Load the dataset
```{r}
getwd()
list.files()
file.exists("UniversalBank.csv")
```


```{r}
getwd()
file.exists("UniversalBank.csv")

```


```{r}
bank <- read.csv("C:/Users/Manager/OneDrive/Documents/DataSets/UniversalBank.csv")


head(bank)
dim(bank)
names(bank)
str(bank)

```

```{r}
table(bank$Personal.Loan)

```
#Create Dummy Variables
```{r}
bank$Education <- factor(bank$Education)
education_dummies <- dummyVars(~ Education, data = bank)

education_data <- predict(education_dummies, newdata = bank)
education_data <- as.data.frame(education_data)

head(education_data)

```

```{r}
names(education_data) <- c("Education_1", "Education_2", "Education_3")

```

#Create the predictor dataset

```{r}
X <- bank %>%
  select(
    Age,
    Experience,
    Income,
    Family,
    CCAvg,
    Mortgage,
    Securities.Account,
    CD.Account,
    Online,
    CreditCard
  )


#Adding dummy variable

X <- cbind(X, education_data)
head(X)

```

#Create the target variable
```{r}
y <- factor(bank$Personal.Loan, levels = c(0, 1))
table(y)

```

#Split the dataset
```{r}
set.seed(123)

#Create the partition
train_index <- createDataPartition(
  y,
  p = 0.60,
  list = FALSE
)

#Create validation and validation dataset

X_train <- X[train_index, ]
X_valid <- X[-train_index, ]

y_train <- y[train_index]
y_valid <- y[-train_index]


#Check the dataset

nrow(X_train)
nrow(X_valid)

```
```{r}
#Standarize the Predictors

preprocess <- preProcess(
  X_train,
  method = c("center", "scale")
)
X_train_scaled <- predict(preprocess, X_train)
X_valid_scaled <- predict(preprocess, X_valid)

```
#Question 1

#Perform a k-NN classification with all predictors except ID and ZIP code using k = 1.

```{r}
knn_1 <- knn(
  train = X_train_scaled,
  test = X_valid_scaled,
  cl = y_train,
  k = 1
)
table(knn_1)

```
```{r}
#Create the Confusion Matrix

confusionMatrix(
  knn_1,
  y_valid,
  positive = "1"
)

```
```{r}
customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1,
  Education_1 = 0,
  Education_2 = 1,
  Education_3 = 0
)
customer


```

#Scaling
```{r}
X_train_scaled <- scale(X_train)

```

#Scale the validation data using the training mean and standard deviation
```{r}
customer_scaled <- scale(
  customer,
  center = attr(X_train_scaled, "scaled:center"),
  scale = attr(X_train_scaled, "scaled:scale")
)

#Run knn with k=1
customer_k1 <- knn(
  train = X_train_scaled,
  test = customer_scaled,
  cl = y_train,
  k = 1
)

customer_k1

```
# Here k-NN with k = 1, so the customer is classified as 0 (Personal Loan = No). Therefore, the model predicts that this customer will not accept the personal loan offer.

#Question 2

```{r}
#What is a choice of k that balances between overfitting and ignoring the predictor information?

k_values <- seq(1, 25, 2)
accuracy <- c()

for (k in k_values) {
  knn_model <- knn(
    X_train_scaled,
    X_valid_scaled,
    y_train,
    k = k
  )
  
  accuracy <- c(
    accuracy,
    mean(knn_model == y_valid)
  )
}

results <- data.frame(
  k = k_values,
  accuracy = accuracy
)

results

```
```{r}
best_k <- results$k[which.max(results$accuracy)]
best_k
results

```
# k = 1 because it produced the highest validation accuracy of 96.45% among the k values tested. However, k = 1 can be sensitive to individual observations and may overfit the training data.

#Question 3
#Show the confusion matrix for the validation data that results from using the best k.

```{r}
best_prediction <- knn(
  X_train_scaled,
  X_valid_scaled,
  y_train,
  k = best_k
)
table(
  Predicted = best_prediction,
  Actual = y_valid
)

```

#Calculate the Accuracy
```{r}
(1793 + 136) / (1793 + 56 + 15 + 136)

```
#The validation confusion matrix shows 1,793 correctly classified customers in class 0 and 136 correctly classified customers in class. The validation accuracy was 96.45%. The sensitivity for the loan-acceptance class was 70.83%, while specificity was 99.17%.


#Question 4
# Classify the customer using the best k

```{r}
customer_best <- knn(
  X_train_scaled,
  customer_scaled,
  y_train,
  k = best_k
)

customer_best
```

#The selected value of k = 1, the customer is classified as 0 (Personal Loan = No).So,the model predicts that the customer will not accept the personal loan.

#Question 5
#Compare the confusion matrix of the test set with that of the training and validation sets. Comment on the differences and their reason.

```{r}
set.seed(123)

train_index <- createDataPartition(
  y,
  p = 0.50,
  list = FALSE
)
```


```{r}
X_train2 <- X[train_index, ]
y_train2 <- y[train_index]

X_remaining <- X[-train_index, ]
y_remaining <- y[-train_index]


```

#Split the data
```{r}
valid_index <- createDataPartition(
  y_remaining,
  p = 0.60,
  list = FALSE
)
```

#Create Validation and testing
```{r}
X_valid2 <- X_remaining[valid_index, ]
y_valid2 <- y_remaining[valid_index]

X_test2 <- X_remaining[-valid_index, ]
y_test2 <- y_remaining[-valid_index]

nrow(X_train2)
nrow(X_valid2)
nrow(X_test2)

```

#Scaling
```{r}
X_train2_scaled <- scale(X_train2)

X_valid2_scaled <- scale(
  X_valid2,
  center = attr(X_train2_scaled, "scaled:center"),
  scale = attr(X_train2_scaled, "scaled:scale")
)

X_test2_scaled <- scale(
  X_test2,
  center = attr(X_train2_scaled, "scaled:center"),
  scale = attr(X_train2_scaled, "scaled:scale")
)

```

#Best k
```{r}
pred_train <- knn(
  X_train2_scaled,
  X_train2_scaled,
  y_train2,
  k = best_k
)

pred_valid <- knn(
  X_train2_scaled,
  X_valid2_scaled,
  y_train2,
  k = best_k
)

pred_test <- knn(
  X_train2_scaled,
  X_test2_scaled,
  y_train2,
  k = best_k
)

```

#Confusion Matrix
```{r}
#Training
table(
  Predicted = pred_train,
  Actual = y_train2
)

```

#Validation
```{r}
table(
  Predicted = pred_valid,
  Actual = y_valid2
)


```

#Test
```{r}
table(
  Predicted = pred_test,
  Actual = y_test2
)

```


#Training: 100% accuracy
#Validation: 96.67% accuracy
#Test: 96.70% accuracy.

#Difference is overfitting caused by k = 1 and the training performance is higher than the validation and test performance because k-NN with k = 1 can overfit the training data.
